### 2023-12-12 manuscript.tex
Editorial purpose (#9) [ci skip]
This build is based on
https://github.com/MMV-Lab/im2im-paper/commit/2dd6a10f7098d9d0ba158bfa18b15f0c218196e6.

This commit was created by the following CI build and job:
https://github.com/MMV-Lab/im2im-paper/commit/2dd6a10f7098d9d0ba158bfa18b15f0c218196e6/checks
https://github.com/MMV-Lab/im2im-paper/actions/runs/7181467740

This manuscript
(\href{https://MMV-Lab.github.io/im2im-paper/v/a66b2deaac1df041285ed3fcaad6976564efd38f/}{permalink})
was automatically generated
from \href{https://github.com/MMV-Lab/im2im-paper/tree/a66b2deaac1df041285ed3fcaad6976564efd38f}{MMV-Lab/im2im-paper@a66b2de}
on October 27, 2023.

@@

Over the past decade, deep learning (DL) research in computer vision has been growing rapidly, with many advances in DL-based image analysis methods for biomedical problems. In this work, we introduce \emph{MMV\_Im2Im}, a new open-source python package for image-to-image transformation in bioimaging applications. \emph{MMV\_Im2Im} is designed with a generic image-to-image transformation framework that can be used for a wide range of tasks, including semantic segmentation, instance segmentation, image restoration, and image generation, etc.. Our implementation takes advantage of state-of-the-art machine learning engineering techniques, allowing researchers to focus on their research without worrying about engineering details. We demonstrate the effectiveness of \emph{MMV\_Im2Im} on more than ten different biomedical problems, showcasing its general potentials and applicabilities.

@@

Our package \emph{MMV\_Im2Im} employs pytorch-lightning {[}\protect\hyperlink{ref-YbvSvdyB}{14}{]} as the core in the backend, which offers numerous benefits, such as readability, flexibility, simplicity and reusability. First of all, have you ever had the moment when you wanted to extend someone's open-source code to suit your special ML needs, but found it so difficult to figure out where and how to extend, especially for complex methods? Or, have you ever encountered the situation where you want to compare the methods and code from two different papers, even solving the same problem, e.g.~semantic segmentation, but not quite easy to grasp quickly since the two repositories are implemented in very different ways? It is not rare that even different researchers from the same group may implement similar methods in very different manners. This is not only a barrier for other people to learn and re-use the open-source code, but also poses challenges for developers in maintenance, further development, and interoperability among different packages. We follow the pytorch-lightning framework and carefully design a universal boilerplate for image-to-image transformation for biomedical applications, where the implementation of all the methods share the same modularized code structure. For all PyTorch users, this greatly lowers the barrier for people to read and understand the code, and makes implementing new methods or extending existing methods simple and fast, at least from an engineering perspective.

@@

The toolbox is designed for both people with or without extensive experience with ML and Python programming. It is not rare to find biomedical image analysis software that is very easy to use on a set of problems, but very hard to extend or adjust to other different but essentially related problems, or find some with great flexibility in tunable knobs at all levels, but unfortunately not easy for inexperienced users. To address this issue, we design the toolbox in a systematically modularized way with various levels of configurability. One can use the toolbox with a single command as simple as \texttt{run\_im2im\ -\/-config\ train\_semanticseg\_3d\ -\/-data.data\_path\ /path/to/data} or make customization on details directly from a human-readable configuration file, such as choosing batch normalization or instance normalization in certain layers of the model, or adding extra data augmentation steps, etc.. We provide an extensive list of more than 20 example configurations for various applications and comprehensive documentation to address common questions for users as reference. For users preferring graphical interface, another napari plugin for the MMV toolbox has been planned as the extension of \emph{MMV\_Im2Im} (see Discussion for details).

@@

\emph{Bottleneck: Not enough consideration for specific chanllengs in microscopy images in general DL toolboxes}

The original idea of a general tool box actually stemmed from the OpenMMLab project (\url{https://openmmlab.com/}), which provides generic codebases for a wide range of computer vision research topics. For instance, \emph{MMSegmentation} (\url{https://github.com/open-mmlab/mmsegmentation}) is an open source toolbox for semantic segmentation, supporting unified benchmarking and state-of-the-art models ready to use out-of-box. It has become one of most widely used codebase for research in semantic segmentation (2.3K forks and 6.5K stars on Github as of September 29, 2023). This inspires us to develop \emph{MMV\_Im2Im} to facilitate research in image-to-image transformation with a special focus on biomedical applications.

First of all, different from general computer vision datasets, such as ImageNet {[}\protect\hyperlink{ref-lt4BNUoG}{16}{]}, where the images are usually small 2D RGB images (e.g., 3 x 256 x 256 pixels), biomedical applications usually involves large-scale high dimensional data (e.g., 500 images of 4 x 128 x 2048 x 2048 voxels). To deal with this issue, we employ the PersistentDataset in MONAI {[}\protect\hyperlink{ref-UU62HYC6}{17}{]} with partial loading and sampling support, as well as delayed image reading powered by aicsimageio {[}\protect\hyperlink{ref-gsfWGJKf}{18}{]} as default (configurable if another dataloader is preferred). As a result, in our stress test, training a 3D nuclei instance segmentation model with more than 125,000 3D images can be conducted efficiently in a day, even with limited resources.

@@

Third, the toolbox pre-packs common functionalities specific to microscopy images. For example, we incorporate the special image normalization method introduced in {[}\protect\hyperlink{ref-Yq8wZ6hc}{4}{]}, where only the middle chunk along the Z dimension of 3D microscopy images will be used for calculating the mean and standard deviation of image intensity for standard normalization. Also, 3D light microscopy images are usually anisotropic, i.e., much lower resolution along Z than XY dimension. So, we adopt the anisotropic variation of UNet as proposed in {[}\protect\hyperlink{ref-jM3v1UjQ}{19}{]}.

Finally, to deploy the model in production, a model trained on small 3D patches sometimes needs to be applied on much larger images. Combining the efficient data handling of aicsimageio {[}\protect\hyperlink{ref-gsfWGJKf}{18}{]} and the sliding window inference with Gaussian weighted blending, the toolbox can yield efficient inference without significant stitching artifacts in production.

All in all, the \emph{MMV\_Im2Im} toolbox stands on the shoulders of many giants in the open-source software and ML engineering communities (pytorch-lightning, MONAI, aicsimageio, etc.) and is systematically designed for image-to-image transformation R\&D for biomedical applications. The source code of \emph{MMV\_Im2Im} is available at \url{https://github.com/MMV-Lab/mmv_im2im}. This manuscript is generated with the open-source package Manubot {[}\protect\hyperlink{ref-YuJbg3zO}{20}{]}. The manuscript source code is available at \url{https://github.com/MMV-Lab/im2im-paper}.

@@

It should be noted that while we recognize the importance of systematically evaluating the predictions, such an analysis falls outside the scope of this paper. We argue that an appropriate evaluation methodology should depend on specific downstream quantitative analysis goals (e.g., {[}\protect\hyperlink{ref-C2iqR6xE}{23}{]}). For example, if our aim is to quantify the size of nucleoli, we must compare the segmentation derived from real nucleoli signals to that of the predicted nucleoli segmentation, ensuring that measurements from both are consistent. Alternatively, if the goal is to localize the nucleoli roughly within the cell, Pearson correlation may be a more appropriate metric. In this work, we concentrate on visual inspection, using Pearson correlation and structural similarity as a rough quantitative reference. Our intent is to demonstrate the utility of our \emph{MMV\_Im2Im} package, and leave appropriate evaluations to users in their specific problems in real studies.

\emph{2D Labelfree:} We started with a simple problem using 2D images from the HeLa ``Kyoto'' cells dataset {[}\protect\hyperlink{ref-xv2VIyRP}{24}{]}. For all images, we took the brightfield channel and the mCherry-H2B channel out of the multi-channel timelapse movies. 2D images were acquired at 20x with 0.8 N.A. and then downscaled by 4 (pixel size: 0.299 nm x 0.299 nm). Example predictions can be found in Figure \ref{fig:labelfree_comparison}-A. We compared a basic UNet model {[}\protect\hyperlink{ref-TutLhFSz}{9}{]} and a 2D version of the fnet model in {[}\protect\hyperlink{ref-Yq8wZ6hc}{4}{]}. The fnet model achieved slightly more accurate predictions than the basic UNet, as seen in Figure \ref{fig:labelfree_comparison}-A.

\emph{3D Labelfree:} We tested with 3D images from the hiPS single cell image dataset {[}\protect\hyperlink{ref-5sGcmDuy}{25}{]}. Specifically, we extracted the brightfield channel and the structure channel from the full field-of-view (FOV) multi-channel images, from the HIST1H2BJ, FBL, NPM1, LMNB1 cell lines, so as to predict from one brightfield image various nuclear structures, histones, nucleoli (dense fibrillar component via fibrillarin), nucleoli (granular component via nucleophosmin), and nuclear envelope, respectively. Images were acquired at 100x with 1.25 NA (voxel size: 0.108 micron x 0.108 micron x 0.29 micron).

@@

Second, we compared different network backbone architectures, including the original fnet model {[}\protect\hyperlink{ref-Yq8wZ6hc}{4}{]}, an enhanced UNet {[}\protect\hyperlink{ref-M7480NLD}{26}{]}, the attention UNet {[}\protect\hyperlink{ref-OCow1hly}{27}{]}, two transformer-based models, SwinUNETR {[}\protect\hyperlink{ref-ZWL3IrVc}{28}{]} and UNETR{[}\protect\hyperlink{ref-XCKUntOB}{29}{]} (all with center normalization). Inspecting the predictions on a holdout validation set suggested that fnet achieved the best performance.

Finally, we showed the comparison between three different types of models, an FCN-type model (i.e., fnet), a pix2pix-type model, and a cycleGAN-type model. For fair comparison, we used fnet as the same backbone for all three types of models. In theory, the pix2pix-type model can be trained in two different ways: from scratch or initializing the generator with a pre-trained fnet (trained as FCN). Examples of the comparison results were shown in the last row in Figure \ref{fig:labelfree_all_structures}-B. Visually, it is evident that the additional adversarial components (i.e., the discriminator) could generate images with more realistic appearance than a typical FCN-type model alone, but again, we leave appropriate evaluations to users' specific biomedical studies.

@@

Segmentation is a common image processing task, and can be considered as a special type of image-to-image transformation, where the generated images are segmentation masks. DL-based methods have achieved huge success in semantic segmentation in biomedical images. In this example, we demonstrated \emph{MMV\_Im2Im} on a pathology application to segment glands from hematoxylin and eosin (H\&E) stained tissue images from the 2015 Gland Segmentation challenge {[}\protect\hyperlink{ref-45Sirz1X}{30},\protect\hyperlink{ref-XAffSYIR}{31}{]}. Stain normalization is an important pre-processing step in order to develop models robust to stain variation and tissue variations. \emph{MMV\_Im2Im} included a classic stain normalization method {[}\protect\hyperlink{ref-tQhnZyjK}{32}{]} as a pre-processing step. The effect of stain normalization can be observed in Figure \ref{fig:2d_gland}-A and B. We trained a simple attention UNet model {[}\protect\hyperlink{ref-OCow1hly}{27}{]}. Evaluated on the two different holdout test sets, the model achieved F1-score, 0.904 ± 0.060 and 0.861 ± 0.117 on test set A and test set B, respectively. The performance was competitive compared to the methods reported in the challenge report {[}\protect\hyperlink{ref-XAffSYIR}{31}{]}, especially with much more consistent performance across the two different test sets. Example results can be found in Figure \ref{fig:2d_gland}-C.

@@

Instance segmentation is a type of segmentation problem that goes beyond semantic segmentation. The goal is to differentiate not only between different types of objects, but also different instances of the same type of objects. Currently, the \emph{MMV\_Im2Im} package supports EmbedSeg-type models. The major benefit of EmbedSeg-type models is their agnosticism to the morphology and dimensionality of the object instances, compared to other models such as StarDist {[}\protect\hyperlink{ref-tIIG2f8K}{33},\protect\hyperlink{ref-14h90Vfg0}{34}{]}, SplineDist {[}\protect\hyperlink{ref-17Yrl6WGQ}{35}{]} and Cellpose {[}\protect\hyperlink{ref-TugPkOLy}{36}{]}. For example, different from the others, EmbedSeg-type models are even able to generate instance segmentation where each instance contains multiple connected components. Additional frameworks such as Omnipose {[}\protect\hyperlink{ref-lXzmjM5n}{37}{]} will be supported in future versions. Another mainstream category of instance segmentation methods are detection-based models, such as Mask-RCNN {[}\protect\hyperlink{ref-xi8wnibR}{38}{]}. However, these models are better suited to the detection framework rather than image-to-image transformation (see Discussion section for details).

The \emph{EmbedSeg}-type models were re-implemented according to the original paper {[}\protect\hyperlink{ref-K2ugNcVa}{10},\protect\hyperlink{ref-QmYuUQ5K}{11}{]} following the generic boilerplate in \emph{MMV\_Im2Im}, with significant improvement. First of all, following the modular design of \emph{MMV\_Im2Im}, it is flexible to use different neural network models as the backbone. For 3D anisotropic microscopy images, the original backbone ERFNet {[}\protect\hyperlink{ref-XAkgs3Nh}{39}{]} doesn't take the anisotropic dimensions into account and therefore may not perform well or even be applicable. In this scenario, it is straightforward to employ another anisotropic neural network bone, such as the anisotropic U-Net in {[}\protect\hyperlink{ref-jM3v1UjQ}{19}{]} or the anisotropic version of Dynamic U-Net in MONAI. Second, we significantly improve training strategy. The original version requires pre-cropping patches centered around each instance and pre-calculated the center images and class images. This may generate a massive amount of additional data on the disk. More importantly, such pre-cropping makes data augmentation nearly impossible, except the simple ones like flipping (otherwise, the pre-calculated centers might be wrong), and also greatly undersamples around negative cases (e.g., background). For example, we have observed that for an EmbedSeg model training only with patches centered around instances, the model may suffer from degraded performance during inference when there are a large amount of background areas without any instances. Again, following the modular design of \emph{MMV\_Im2Im}, it is now possible to do on-the-fly data augmentation and patch sampling, even weighted patch sampling. Another extension compared to the original implementation was that the \emph{MMV\_Im2Im} package made sliding windowing inference straightforward, and therefore permitted easy handling of images of any size during inference.

In this work, we tested on both 2D and 3D instance segmentation problems. Going from 2D to 3D is not a simple generalization from 2D models by switching 2D operations with 3D operations, but with many practical challenges. Large GPU footprint is one of the biggest issues, which makes many training strategies common in 2D not feasible in 3D, e.g.~limited mini-batch size. \emph{MMV\_Im2Im} is able to take advantage of state-of-the-art ML engineering methods to efficiently handle 3D problems. For example, by using effective half-precision training, one can greatly reduce GPU memory workload for each sample and therefore increase the batch size or the patch size. When multiple GPUs are available, it is also possible to easily take advantage of the additional resources to scale up the training to multiple GPU cards, even multiple GPU nodes. As a demonstration, we applied \emph{EmbedSeg}-like models to a 2D problem of segmenting \emph{C. elegans} from widefield images {[}\protect\hyperlink{ref-138foKNOh}{40}{]}, as well as a 3D problem of nuclear segmentation from fluorescent and brightfield images from the hiPS single cell image dataset {[}\protect\hyperlink{ref-5sGcmDuy}{25}{]}.

For the 2D problem, we adopted the same network backbone as in the original \emph{EmbedSeg} paper. Example results on a small holdout set of 5 images are shown in Figure \ref{fig:instance}-A (average precision at 50 = 0.866 ± 0.163), which is comparable to the original published results {[}\protect\hyperlink{ref-QmYuUQ5K}{11}{]}. For the 3D problem, the original backbone is not directly applicable, due to the aforementioned anisotropic issue and the images in the dataset do not contain enough Z-slices to run through all down sampling blocks in 3D. The anisotropic UNet {[}\protect\hyperlink{ref-jM3v1UjQ}{19}{]} is used here.

@@

We did a special comparison in this subsection to further illustrate the difference between semantic and instance segmentations. We took the 3D fibrillarin dataset from {[}\protect\hyperlink{ref-5sGcmDuy}{25}{]}. There are multiple channels in each 3D image, including DNA dye, membrane dye, and the structure channel (i.e., fibrillarin). The original fibrillarin segmentation released with the dataset is binary semantic segmentation. With additional cell segmentations available in the dataset, we can know which groups of segmented fibrillarin belong to the same cell. Then, we can convert the semantic segmentation into 3D instance segmentation ground truth (fibrillarin pixels belonging to the same cell are grouped as a unique instance). Sample images and results are shown in Figure \ref{fig:3dseg}. We can observe that the semantic segmentation model is able to achieve good accuracy in determining pixels from the fibrillarin signals (F1 = 0.958 ± 0.008). Meanwhile, the instance segmentation can group them properly (average precision at 50 = 0.795 ± 0.055) so that fibrillarin masks from the same cell are successfully identified as unique instances, even without referring to cell segmentation results. This is not a simple distance-based grouping, since the fibrillarin signals from tightly touching nuclei may exist close to each other.

@@

To perform unsupervised learning, we used raw images from the hiPS single cell image dataset {[}\protect\hyperlink{ref-5sGcmDuy}{25}{]}, as well as their corresponding segmentations (may not be absolute pixel-wise ground truth, but have gone through systematic evaluation to ensure the overall quality). We shuffled the raw images and their segmentations to generate a set of simulated segmentation masks. A demonstration of the concept is illustrated in Figure \ref{fig:unsupervised}-A. Example results for all 3D models are shown in Figure \ref{fig:unsupervised}-B, and the F1-scores on the test set are summarized in Table \ref{tbl:unsuper}.

>>>>

This manuscript
(\href{https://MMV-Lab.github.io/im2im-paper/v/2dd6a10f7098d9d0ba158bfa18b15f0c218196e6/}{permalink})
was automatically generated
from \href{https://github.com/MMV-Lab/im2im-paper/tree/2dd6a10f7098d9d0ba158bfa18b15f0c218196e6}{MMV-Lab/im2im-paper@2dd6a10}
on December 12, 2023.

@@

Over the past decade, deep learning (DL) research in computer vision has been growing rapidly, with many advances in DL-based image analysis methods for biomedical problems. In this work, we introduce \emph{MMV\_Im2Im}, a new open-source Python package for image-to-image transformation in bioimaging applications. \emph{MMV\_Im2Im} is designed with a generic image-to-image transformation framework that can be used for a wide range of tasks, including semantic segmentation, instance segmentation, image restoration, and image generation, etc.. Our implementation takes advantage of state-of-the-art machine learning engineering techniques, allowing researchers to focus on their research without worrying about engineering details. We demonstrate the effectiveness of \emph{MMV\_Im2Im} on more than ten different biomedical problems, showcasing its general potentials and applicabilities.

@@

Our package \emph{MMV\_Im2Im} employs pytorch-lightning {[}\protect\hyperlink{ref-YbvSvdyB}{14}{]} as the core in the backend, which offers numerous benefits, such as readability, flexibility, simplicity and reusability. First of all, have you ever had the moment when you wanted to extend someone's open-source code to suit your special ML needs, but found it so difficult to figure out where and how to extend, especially for complex methods? Or, have you ever encountered the situation where you want to compare the methods and code from two different papers, even solving the same problem, e.g.~semantic segmentation, but not quite easy to grasp quickly since the two repositories are implemented in very different ways? It is not rare that even different researchers from the same group may implement similar methods in very different manners. This is not only a barrier for other people to learn and re-use the open-source code, but also poses challenges for developers in maintenance, further development, and interoperability among different packages. We follow the pytorch-lightning framework and carefully design a universal boilerplate for image-to-image transformation for biomedical applications, where the implementation of all the methods share the same modularized code structure. For all PyTorch users, this greatly lowers the learning curve for people to read and understand the code, and makes implementing new methods or extending existing methods simple and fast, at least from an engineering perspective.

@@

The toolbox is designed for both people with or without extensive experience with ML and Python programming. It is not rare to find biomedical image analysis software that is very easy to use on a set of problems, but very hard to extend or adjust to other different but essentially related problems, or find some with great flexibility with tunable knobs at all levels, but unfortunately not easy for inexperienced users. To address this issue, we design the toolbox in a systematically modularized way with various levels of configurability. One can use the toolbox with a single command as simple as \texttt{run\_im2im\ -\/-config\ train\_semanticseg\_3d\ -\/-data.data\_path\ /path/to/data} or make customization on details directly from a human-readable configuration file, such as choosing batch normalization or instance normalization in certain layers of the model, or adding extra data augmentation steps, etc.. We provide an extensive list of more than 20 example configurations for various applications and comprehensive documentation to address common questions for users as reference. For users preferring graphical interface, another napari plugin for the MMV toolbox has been planned as the extension of \emph{MMV\_Im2Im} (see Discussion for details).

@@

\emph{Bottleneck: Not enough consideration for specific challenges in microscopy images in general DL toolboxes}

The original idea of a general toolbox actually stemmed from the OpenMMLab project (\url{https://openmmlab.com/}), which provides generic codebases for a wide range of computer vision research topics. For instance, \emph{MMSegmentation} {[}\protect\hyperlink{ref-EcizztLg}{16}{]} is an open source toolbox for semantic segmentation, supporting unified benchmarking and state-of-the-art models ready to use out-of-box. It has become one of most widely used codebase for research in semantic segmentation (2.3K forks and 6.5K stars on Github as of September 29, 2023). This inspires us to develop \emph{MMV\_Im2Im} to facilitate research in image-to-image transformation with a special focus on biomedical applications.

First of all, different from general computer vision datasets, such as ImageNet {[}\protect\hyperlink{ref-lt4BNUoG}{17}{]}, where the images are usually small 2D RGB images (e.g., 3 x 256 x 256 pixels), biomedical applications usually involves large-scale high dimensional data (e.g., 500 images of 4 x 128 x 2048 x 2048 voxels). To deal with this issue, we employ the PersistentDataset in MONAI {[}\protect\hyperlink{ref-UU62HYC6}{18}{]} with partial loading and sampling support, as well as delayed image reading powered by aicsimageio {[}\protect\hyperlink{ref-gsfWGJKf}{19}{]} as default (configurable if another dataloader is preferred). As a result, in our stress test, training a 3D nuclei instance segmentation model with more than 125,000 3D images can be conducted efficiently in a day, even with limited resources.

@@

Third, the toolbox pre-packs common functionalities specific to microscopy images. For example, we incorporate the special image normalization method introduced in {[}\protect\hyperlink{ref-Yq8wZ6hc}{4}{]}, where only the middle chunk along the Z dimension of 3D microscopy images will be used for calculating the mean and standard deviation of image intensity for standard normalization. Also, 3D light microscopy images are usually anisotropic, i.e., much lower resolution along Z than XY dimension. So, we adopt the anisotropic variation of UNet as proposed in {[}\protect\hyperlink{ref-jM3v1UjQ}{20}{]}.

Finally, to deploy the model in production, a model trained on small 3D patches sometimes needs to be applied not only on much larger images. Combining the efficient data handling of aicsimageio {[}\protect\hyperlink{ref-gsfWGJKf}{19}{]} and the sliding window inference with gaussian weighted blending, the toolbox can yield efficient inference without visible stitching artifacts in production.

All in all, the \emph{MMV\_Im2Im} toolbox stands on the shoulders of many giants in the open-source software and ML engineering communities (pytorch-lightning, MONAI, aicsimageio, etc.) and is systematically designed for image-to-image transformation R\&D for biomedical applications. The source code of \emph{MMV\_Im2Im} is available at \url{https://github.com/MMV-Lab/mmv_im2im}. This manuscript is generated with the open-source package Manubot {[}\protect\hyperlink{ref-YuJbg3zO}{21}{]}. The manuscript source code is available at \url{https://github.com/MMV-Lab/im2im-paper}.

@@

It should be noted that while we recognize the importance of systematically evaluating the predictions, such an analysis falls outside the scope of this paper. We argue that an appropriate evaluation methodology should depend on specific downstream quantitative analysis goals (e.g., {[}\protect\hyperlink{ref-C2iqR6xE}{24}{]}). For example, if our aim is to quantify the size of nucleoli, we must compare the segmentation derived from real nucleoli signals to that of the predicted nucleoli segmentation, ensuring that measurements from both are consistent. Alternatively, if the goal is to localize the nucleoli roughly within the cell, Pearson correlation may be a more appropriate metric. In this work, we concentrate on visual inspection, using Pearson correlation and structural similarity as a rough quantitative reference. Our intent is to demonstrate the utility of our \emph{MMV\_Im2Im} package, and leave appropriate evaluations to users in their specific problems in real studies.

\emph{2D Labelfree:} We started with a simple problem using 2D images from the HeLa ``Kyoto'' cells dataset {[}\protect\hyperlink{ref-xv2VIyRP}{25}{]}. For all images, we took the brightfield channel and the mCherry-H2B channel out of the multi-channel timelapse movies. 2D images were acquired at 20x with 0.8 N.A. and then downscaled by 4 (pixel size: 0.299 nm x 0.299 nm). Example predictions can be found in Figure \ref{fig:labelfree_comparison}-A. We compared a basic UNet model {[}\protect\hyperlink{ref-TutLhFSz}{9}{]} and a 2D version of the fnet model in {[}\protect\hyperlink{ref-Yq8wZ6hc}{4}{]}. The fnet model achieved slightly more accurate predictions than the basic UNet, as seen in Figure \ref{fig:labelfree_comparison}-A.

\emph{3D Labelfree:} We tested with 3D images from the hiPS single cell image dataset {[}\protect\hyperlink{ref-5sGcmDuy}{26}{]}. Specifically, we extracted the brightfield channel and the structure channel from the full field-of-view (FOV) multi-channel images, from the HIST1H2BJ, FBL, NPM1, LMNB1 cell lines, so as to predict from one brightfield image various nuclear structures, histones, nucleoli (dense fibrillar component via fibrillarin), nucleoli (granular component via nucleophosmin), and nuclear envelope, respectively. Images were acquired at 100x with 1.25 NA (voxel size: 0.108 micron x 0.108 micron x 0.29 micron).

@@

Second, we compared different network backbone architectures, including the original fnet model {[}\protect\hyperlink{ref-Yq8wZ6hc}{4}{]}, an enhanced UNet {[}\protect\hyperlink{ref-M7480NLD}{27}{]}, the attention UNet {[}\protect\hyperlink{ref-OCow1hly}{28}{]}, two transformer-based models, SwinUNETR {[}\protect\hyperlink{ref-ZWL3IrVc}{29}{]} and UNETR{[}\protect\hyperlink{ref-XCKUntOB}{30}{]} (all with center normalization). Inspecting the predictions on a holdout validation set suggested that fnet achieved the best performance.

Finally, we showed the comparison between three different types of models, an FCN-type model (i.e., fnet), a pix2pix-type model, and a cycleGAN-type model. For fair comparison, we used fnet as the same backbone for all three types of models. In theory, the pix2pix-type model can be trained in two different ways: from scratch or initializing the generator with a pre-trained fnet (trained as FCN). Examples of the comparison results were shown in the last two rows in Figure \ref{fig:labelfree_all_structures}-B. Visually, it is evident that the additional adversarial components (i.e., the discriminator) could generate images with more realistic appearance than a typical FCN-type model alone, but again, we leave the appropriate quantitative evaluations to users' specific biomedical studies.

@@

Segmentation is a common image processing task, and can be considered as a special type of image-to-image transformation, where the generated images are segmentation masks. DL-based methods have achieved huge success in semantic segmentation in biomedical images. In this example, we demonstrated \emph{MMV\_Im2Im} on a pathology application to segment glands from hematoxylin and eosin (H\&E) stained tissue images from the 2015 Gland Segmentation challenge {[}\protect\hyperlink{ref-45Sirz1X}{31},\protect\hyperlink{ref-XAffSYIR}{32}{]}. Stain normalization is an important pre-processing step in order to develop models robust to stain variation and tissue variations. \emph{MMV\_Im2Im} included a classic stain normalization method {[}\protect\hyperlink{ref-tQhnZyjK}{33}{]} as a pre-processing step. The effect of stain normalization can be observed in Figure \ref{fig:2d_gland}-A and B. We trained a simple attention UNet model {[}\protect\hyperlink{ref-OCow1hly}{28}{]}. Evaluated on the two different holdout test sets, the model achieved F1-score, 0.904 ± 0.060 and 0.861 ± 0.117 on test set A and test set B, respectively. The performance was competitive compared to the methods reported in the challenge report {[}\protect\hyperlink{ref-XAffSYIR}{32}{]}, especially with much more consistent performance across the two different test sets. Example results can be found in Figure \ref{fig:2d_gland}-C.

@@

Instance segmentation is a type of segmentation problem that goes beyond semantic segmentation. The goal is to differentiate not only between different types of objects, but also different instances of the same type of objects. Currently, the \emph{MMV\_Im2Im} package supports EmbedSeg-type models. The major benefit of EmbedSeg-type models is their agnosticism to the morphology and dimensionality of the object instances, compared to other models such as StarDist {[}\protect\hyperlink{ref-tIIG2f8K}{34},\protect\hyperlink{ref-14h90Vfg0}{35}{]}, SplineDist {[}\protect\hyperlink{ref-17Yrl6WGQ}{36}{]} and Cellpose {[}\protect\hyperlink{ref-TugPkOLy}{37}{]}. For example, different from the others, EmbedSeg-type models are even able to generate instance segmentation where each instance contains multiple connected components. Additional frameworks such as Omnipose {[}\protect\hyperlink{ref-lXzmjM5n}{38}{]} will be supported in future versions. Another mainstream category of instance segmentation methods are detection-based models, such as Mask-RCNN {[}\protect\hyperlink{ref-xi8wnibR}{39}{]}. However, these models are better suited to the detection framework rather than image-to-image transformation (see Discussion section for details).

The \emph{EmbedSeg}-type models were re-implemented according to the original paper {[}\protect\hyperlink{ref-K2ugNcVa}{10},\protect\hyperlink{ref-QmYuUQ5K}{11}{]} following the generic boilerplate in \emph{MMV\_Im2Im}, with significant improvement. First of all, following the modular design of \emph{MMV\_Im2Im}, it is flexible to use different neural network models as the backbone. For 3D anisotropic microscopy images, the original backbone ERFNet {[}\protect\hyperlink{ref-XAkgs3Nh}{40}{]} doesn't take the anisotropic dimensions into account and therefore may not perform well or even be applicable. In this scenario, it is straightforward to employ another anisotropic neural network bone, such as the anisotropic U-Net in {[}\protect\hyperlink{ref-jM3v1UjQ}{20}{]} or the anisotropic version of Dynamic U-Net in MONAI. Second, we significantly improve training strategy. The original version requires pre-cropping patches centered around each instance and pre-calculated the center images and class images. This may generate a massive amount of additional data on the disk. More importantly, such pre-cropping makes data augmentation nearly impossible, except the simple ones like flipping (otherwise, the pre-calculated centers might be wrong), and also greatly undersamples around negative cases (e.g., background). For example, we have observed that for an EmbedSeg model training only with patches centered around instances, the model may suffer from degraded performance during inference when there are a large amount of background areas without any instances. Again, following the modular design of \emph{MMV\_Im2Im}, it is now possible to do on-the-fly data augmentation and patch sampling, even weighted patch sampling. Third, our improved \emph{EmbedSeg}-type models can accept an exclusion mask so that certain parts of the images can be ignored during training. This is especially useful for partially annotated ground truth. For large images, it could be extremely time-consuming to require every single instance to be annotated. The exclusion masks can address this bottleneck. Another extension compared to the original implementation was that the \emph{MMV\_Im2Im} package made sliding windowing inference straightforward, and therefore permitted easy handling of images of any size during inference.

In this work, we tested on both 2D and 3D instance segmentation problems. Going from 2D to 3D is not a simple generalization from 2D models by switching 2D operations with 3D operations, but with many practical challenges. Large GPU footprint is one of the biggest issues, which makes many training strategies common in 2D not feasible in 3D, e.g.~limited mini-batch size. \emph{MMV\_Im2Im} is able to take advantage of state-of-the-art ML engineering methods to efficiently handle 3D problems. For example, by using effective half-precision training, one can greatly reduce GPU memory workload for each sample and therefore increase the batch size or the patch size. When multiple GPUs are available, it is also possible to easily take advantage of the additional resources to scale up the training to multiple GPU cards, even multiple GPU nodes. As a demonstration, we applied \emph{EmbedSeg}-like models to a 2D problem of segmenting \emph{C. elegans} from widefield images {[}\protect\hyperlink{ref-138foKNOh}{41}{]}, as well as a 3D problem of nuclear segmentation from fluorescent and brightfield images from the hiPS single cell image dataset {[}\protect\hyperlink{ref-5sGcmDuy}{26}{]}.

For the 2D problem, we adopted the same network backbone as in the original \emph{EmbedSeg} paper. Example results on a small holdout set of 5 images are shown in Figure \ref{fig:instance}-A (average precision at 50 = 0.866 ± 0.163), which is comparable to the original published results {[}\protect\hyperlink{ref-QmYuUQ5K}{11}{]}. For the 3D problem, the original backbone is not directly applicable, due to the before mentioned anisotropic issue and the images in the dataset do not contain enough Z-slices to run through all down sampling blocks in 3D. The anisotropic UNet {[}\protect\hyperlink{ref-jM3v1UjQ}{20}{]} is used here. The segmentation results obtained from the public dataset {[}\protect\hyperlink{ref-5sGcmDuy}{26}{]} contain nuclear instance segmentation of all cells. But, the cells touching the image borders are ignored from downstream analysis {[}\protect\hyperlink{ref-5sGcmDuy}{26}{]} and therefore not curated. In other words, the segmentation from this public dataset can only be used as high-quality nuclear instance segmentation ground truth after excluding the areas covered by cells touching the image borders {[}\protect\hyperlink{ref-5sGcmDuy}{26}{]}. Therefore, the exclusion masking function in MMV\_Im2Im is very helpful in this example.

@@

We did a special comparison in this subsection to further illustrate the difference between semantic and instance segmentations. We took the 3D fibrillarin dataset from {[}\protect\hyperlink{ref-5sGcmDuy}{26}{]}. There are multiple channels in each 3D image, including DNA dye, membrane dye, and the structure channel (i.e., fibrillarin in this case). The original fibrillarin segmentation released with the dataset is a semantic segmentation (0=background, 1=fibrillarin). With the additional cell segmentation available in the dataset, we can know which groups of segmented fibrillarin belong to the same cell. Then, we can convert the original 3D fibrillarin semantic segmentation ground truth into 3D instance segmentation ground truth (fibrillarin pixels belonging to the same cell are grouped as a unique instance). Sample images and results are shown in Figure \ref{fig:3dseg}. We can observe that the semantic segmentation model is able to achieve good accuracy in determining pixels from the fibrillarin signals (F1 = 0.958 ± 0.008). Meanwhile, the instance segmentation can group them properly (average precision at 50 = 0.795 ± 0.055) so that fibrillarin masks from the same cell are successfully identified as unique instances, even without referring to the cell membrane channel or cell segmentation results. This is not a simple grouping step based on distance, since the fibrillarin signals from tightly touching nuclei may exist very close to each other.

@@

To perform unsupervised learning, we used raw images from the hiPS single-cell image dataset {[}\protect\hyperlink{ref-5sGcmDuy}{26}{]}, as well as their corresponding segmentations (may not be absolute pixel-wise ground truth, but have gone through systematic evaluation to ensure the overall quality). We shuffled the raw images and their segmentations to generate a set of simulated segmentation masks. A demonstration of the concept is illustrated in Figure \ref{fig:unsupervised}-A. Example results for all 3D models are shown in Figure \ref{fig:unsupervised}-B, and the F1-scores on the test set are summarized in Table \ref{tbl:unsuper}.

For the 2D example, we saw that the unsupervised training provides a valuable segmentation, which is reflected by the F1 Score in Table \ref{tbl:unsuper}. For the 3D examples, it has been suggested that the quality of unsupervised nuclei segmentation could be further improved with additional simulation strategies {[}\protect\hyperlink{ref-RuFP3CS3}{12}{]}. Overall, we believe that unsupervised learning offers an effective way to generate preliminary segmentation, which can be further refined through active learning such as the iterative DL workflow described in {[}\protect\hyperlink{ref-jM3v1UjQ}{20}{]}.
